HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-27 docs/HLL_exp_bad/main.tex
todo cmd
complete draft

\documentclass{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}

@@

\usepackage{xcolor}
\definecolor{oxygenorange}{HTML}{FFDD00}
\usepackage[color=oxygenorange]{todonotes}
\newcommand{\mathilde}[2][inline]{\todo[#1]{\textbf{Mathilde:} #2}\xspace}
\newcommand{\kenny}[2][inline]{\todo[#1]{\textbf{Kenny:} #2}\xspace}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary can make HLL underestimate true cardinality by a factor that is exponential in the number of buckets used by the algorithm.

@@

\subsection{Adversarial Setting}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, the \textit{insider} setup of \cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since we assume that the attacker owns the data structure and thus has an exhaustive view on the sketch. %combines the knowledge of adversaries from S1 and S2.
\kenny{Not really! Because S1 is actually quite unrealistic, as we will argue below.}

\begin{itemize}
\item[S1:] The adversary has access to a \textit{test oracle} provided by the sketch owner, which allows him to submit an item and learn the hypothetical effect of inserting it into the sketch. He is then able to determine if inserting this item \textit{would} increase the cardinality estimation.

%The adversary is able to test the hypothetical effect of the insertion of an item on the cardinality estimation of the sketch. This can achieved via a \textit{test oracle} provided by the sketch owner, which allows an user to test the effect of an insertion on the cardinality estimate but without actually inserting the item, or if the attacker would gain access to a black-box replica of the sketch in use, i.e., buying the same switch and testing in the lab. %black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item.
\kenny{TODO: clarify more what this is terms of a ''test oracle''; its stronger than described because the adverary can test the effect of an insertion without making the insertion!s}
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. \mathilde{Would it make sense to add a sentence "This scenario would apply to~\cite{portscanhll}."?}
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, he can access and interact with the sketch via an API provided by the sketch owner, allowing him to insert an item or ask for the HLL's cardinality estimate at any point in time.
\kenny{TODO: The external attacker from~\cite{cardestprivacy}  described in terms of oracles.}
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate. Again, an S4 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack, but it can compute these values for itself (assuming its attack proceeds without input from other users being added to the HLL sketch in parallel with the attack). \mathilde{I am not sure about the last sentence. In the \textit{insider} model of \cite{cardestprivacy} the attacker owns the sketch so has access to both statistics (cardinality estimation) and counters in real time. So if he receives data coming from another user / notices an update in the counters, he can check reiterate over the discarded items and check whether he can retain some more? I know the goal of \cite{cardestprivacy} is different and there it makes sense to consider the scenario where the owner is malicious but not necessarily for us.}
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln} and S3 and S4 as the \textit{external} and \textit{internal} attack scenarios in \cite{cardestprivacy}.

We start by highlighting the fact that S1 seems unrealistic in practice, as there is no reason why a sketch owner would provide such \textit{test oracle}. The authors of \cite{hllvuln} argue that the assumptions of S1 could also be fulfilled if the attacker gains access to a replica of the machine the HLL is implemented on. The adversary could run tests on it, sequentially inserting, asking for the cardinality estimation and resetting the device, retaining only the items which do not increment the HLL estimate. \mathilde{Todo: Add sentence This second approach does not look much more convincing/practical.}
%This is unclear what this means in practice, since after insertion into the HLL sketch, an item cannot usually be removed by the adversary.
\kenny{Mathilde: can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more.}

Furthermore, we show that although \cite{cardestprivacy} presents S3 and S4 as detached scenarios, the attacker with the capabilities of S3 can obtain the knowledge of S4. By analyzing which inputs increase the cardinality estimation, an attacker can easily recover the contents of each bucket in the sketch. To guess the estimator of a targeted bucket, the attacker gradually inserts items with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$, until the cardinality estimation of the sketch increases. Once a change occurs, the attacker knows that the maximum number of leading zeros observed in that bucket just got updated to a value that can be recovered by looking at the last inserted item. The attacker can determine all estimators with $O(2^{n + \ell})$ pre-computations and $O(\ell \cdot 2^{n})$ insert and cardinality estimation queries. The fuller the sketch the most work the attacker has to provide.

\kenny{TODO: Show that S3 implies S4 $O(2^{n + \ell})$ precomputation and $O(\ell \cdot 2^{n})$ insert and read queries. So we. focus from now on S2 and S4 -- because S1 is unrealistic and S3 implies S4.}

As the assumptions of S1 appear unrealistic and an attacker can escalate from the knowledge of S3 to the one of S4, we focus on the scenarios S2 and S4. We first present our attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln}, and then use the additional information the adversary has access to in scenario S4 to make the attack even more powerful.

@@

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  2$^{t-1}$ distinct items without increasing the HLL cardinality estimate at all, under the assumption that all buckets in the HLL sketch already hold a value $c_i ge 1$. On the other hand, if some buckets are empty (but the adversary does not know which ones in the blind setting of S2), then we may expect a moderate increase in the HLL cardinality estimator. In the worst case, when all buckets are initially empty, the cardinality estimate will increase from 0 to at most $\alpha2^{n+1}$ (regardless of the number of items the adversary inserts).

@@

%\textit{NOTE: does it take into account the special case of empty bucket ? counters are initialized to $c_i=-oo$ (or -1 in the code???) thus $P(h(x)$ has LESS than $-oo$ leading $0s) = 0$ but it is hard to add it to the probabilities while staying clear.}\\
%2$^{len(h(x))}\times \frac{2H_c-1}{H_c}$ items into the sketch.\\

Note that the above analysis for scenario S4 made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are 0, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert without increasing the HLL cardinality estimate becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$. Of course, the adversary can relax its attack and allow some of the empty buckets to be hit. This increases the insertion rate at the cost of an increased final cardinality estimate.

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. The blind adversary in scenario S2 has no way to detect that it is attacking an empty HLL and cannot adapt its strategy, so all buckets that are hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S4 is able to detect that all buckets are empty and adapt its attack accordingly. The adversary can now target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality-estimate trade-offs are of course possible.

In both scenarios S2 and S4, we see that an adversary who has $t$ bits of flexibility in its inputs can expect to insert exponentially (in $t$) many inputs values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the cases where the HLL sketch has empty buckets in S2 or where the HLL sketch is empty to begin with in S2 and S4). The result is largely independent of the HLL parameters, but does depend on the extent to which the HLL is already filled, via a factor $1-(2H_c)^{-1}$ and the number of empty buckets $r$, via a factor $1- \frac{r}{2^n}$.

@@

As a proof-of-concept, we implemented our attacks against the HLL implementation from \cite{clahll}. The attack code is available at \href{https://github.com/PizzaWhisperer/HLLVuln}{\textit{\url{https://github.com/PizzaWhisperer/HLLVuln}}}. For simplicity, our items are 4-character long ASCII strings, giving us up to 7,311,616 distinct items that can be inserted. We choose $h$ to be the 32-bit Murmur3 hash function from \cite{murmur3code}. \kenny{Why do we get to choose $h$? Isn't it fixed by the specific implementation we are attacking?} \mathilde{The library has a place holder for a 32 bit hash function but lets the user choose it. This was one of my concerns and I was wondering if it would make sense to attack \cite{datasketch} instead (hardcoded $h$, but when we discussed it, you said that due to Kerckhoff's principle, this hash function should be public so it does not matter that much.} We create an HLL sketch using $2^n = 256$ buckets and initialize it with 1,000 random items representing honest users' data.

We then challenge the adversary to pick the largest possible number of items to add to the sketch from our set of. while trying to keep the cardinality estimate as low as possible, either in the S2 or the S4 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.

Those empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the adversary is able to hide approximately half of the items' set under S2, and around 83\%%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$).
$\approx (1-\frac{1}{2\times H_c})$ with $H_c\approx\frac{1'000}{256}=3.9$, under S4. Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by the Eq. (29) of \cite{hll} or 1 in respectively S2 and S4.

To assess the efficiency of our attacks, we can confront our results from Tab. \ref{table:tab2} with the attack presented in the Section 5.1 of \cite{hllvuln}, where they target the Redis \cite{redishll} open-source implementation of HLL and perform the attack against an empty sketch. Given a set of 250'000 items, the adversary could choose ``74'390 distinct items and obtain an HLL estimate of only 15'780``~\cite{hllvuln} achieving a five-fold reduction from the true cardinality. In comparison, the adversary following our strategy is able to add 125'000 distinct items while keeping the estimate at 367, demonstrating a reduction by a factor of 340.

\begin{table}[h]
\caption{Result of the attacks over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Adv. Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'000 & 0 & 1'005 & 0 \\ \hline
    \textbf{\# Items added} & 3'655'744 & 3'655'740 & 6'090'003 & 14'220 \\ \hline
    \textbf{Final Est. Card.} & 1'011 & 367 & 1'005 & 1 \\ \hline
\end{tabular}
\label{table:tab1}
\end{table}

\begin{table}[h]
\caption{Result of the attacks in the setting of \cite{hllvuln} over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Adv. Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'004 & 0 & 998 & 0 \\ \hline
    \textbf{\# Items added} & 124'940 & 125'085 & 204'669 & 501 \\ \hline
    \textbf{Final Est. Card.} & 1'058 & 367 & 998 & 1 \\ \hline
\end{tabular}
\label{table:tab2}
\end{table}

\section{Conclusion}\label{sec:conclusions}
The HyperLogLog algorithm is an elegant and efficient solution to the distinct count problem. Its simple structure makes it easy to code and use, as shows the growing number of available open-source implementations. Nonetheless, disclosing such low level details can be exploited by malicious users to manipulate the estimate of the sketch and break the system that is built on top. Our attack is trivial but powerful and should raise awareness on the security of HLL, such that product designers understand the risks they may encounter when exposed to an adversarial setting. Future work could consist in finding mitigations that do not rely on salt. For example, would the imposed ID setting (like Facebook) solve the issue but still be convenient for the various HLL applications? Also left for future work is to formalize the extension of our attack to the HLL variants, i.e., to show that it also breaks the HLL++ \cite{hllpratice} and sliding HLL \cite{slidinghll} algorithms.

>>>>

\documentclass{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xspace}

@@

\usepackage{xcolor}
\definecolor{oxygenorange}{HTML}{FFDD00}
\usepackage[color=oxygenorange]{todonotes}
\newcommand{\mathilde}[1]{\todo[inline]{\textbf{Mathilde:} #1}\xspace}
\newcommand{\kenny}[1]{\todo[inline]{\textbf{Kenny:} #1}\xspace}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of $O(2^t)$ where $t$ is th number of bits of flexibility in the input).

@@

\subsection{Adversarial Setting}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, the \textit{insider} setup of \cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since it assumes that the attacker has a perfect view on the sketch. %combines the knowledge of adversaries from S1 and S2.
%\kenny{Not really! Because S1 is actually quite unrealistic, as we will argue below.}

\begin{itemize}
\item[S1:] The adversary has access to a \textit{test oracle} provided by the sketch owner, which allows it to submit an item and learn the hypothetical effect of inserting it into the sketch. It is then able to determine if inserting this item \textit{would} increase the cardinality estimate, without actually inserting it. 
%The adversary is able to test the hypothetical effect of the insertion of an item on the cardinality estimation of the sketch. This can achieved via a \textit{test oracle} provided by the sketch owner, which allows an user to test the effect of an insertion on the cardinality estimate but without actually inserting the item, or if the attacker would gain access to a black-box replica of the sketch in use, i.e., buying the same switch and testing in the lab. %black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item.
%\kenny{TODO: clarify more what this is terms of a ''test oracle''; its stronger than described because the adversary can test the effect of an insertion without making the insertion!s}
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}. 
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. 
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound). 
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. S3 is the \textit{external} attack scenario in~\cite{cardestprivacy}. S4 is a slightly weaker version of the \textit{insider} attack scenario in~\cite{cardestprivacy}, the difference being that an insider adversary in~\cite{cardestprivacy} has continuous access to the HLL sketch internals including all bucket values, while our adversary in S4 only has access to a snapshot of the HLL sketch internals at the start of its attack.

We start by highlighting the fact that S1 seems unrealistic in practice, as there is no reason why a typical HLL implementation would provide such a test oracle. The authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker gains access to a replica of the machine on which the HLL is implemented. The adversary could run tests on it, sequentially inserting, asking for the cardinality estimation and resetting the device, retaining only the items which do not increment the HLL estimate. This may be possible under certain circumstances, but we do not consider it typical of HLL deployments.

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once a change occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded.

A version of this attack can be carried out to elevate an external adversary to a full-strength insider adversary as per~\cite{cardestprivacy} (instead of our slightly weaker S4 adversary), since it can simply be done repeatedly each time the external adversary wants to know the complete internals of the HLL sketch. Obvious optimisations can be carried out to reduce the query complexity each time the attack is run (since the bucket contents can only increase over time, some queries are redundant; moreover, the pre-computation is only needed once and is moderate given typical HLL parameters, e.g.\ $n+\ell = 32$).

Since the assumptions of S1 appear to us to be unrealistic and since an attacker can escalate from S3 to S4, we focus on the scenarios S2 and S4 in what follows. We first present an attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln} (in their equivalent scenario M1). We then use the additional information the adversary has access to in scenario S4 to make the attack even more powerful.

@@

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  $2^{t-1}$ distinct items without increasing the HLL cardinality estimate at all, under the assumption that all buckets in the HLL sketch already hold a value $c_i \ge 1$. On the other hand, if some buckets are empty (but the adversary does not know which ones in the blind setting of S2), then we may expect a moderate increase in the HLL cardinality estimator. In the worst case, when all buckets are initially empty, the cardinality estimate will increase from 0 to at most $\alpha2^{n+1}$ (regardless of the number of items the adversary inserts). Notice that in all cases, provided $t$ is large enough, the adversary can keep the cardinality estimate linear in the total number of buckets ($2^n$) while the expected value of the true cardinality, $2^{t-1}$, can be made as large as the adversary pleases.

@@

%\textit{NOTE: does it take into account the special case of empty bucket ? counters are initialised to $c_i=-oo$ (or -1 in the code???) thus $P(h(x)$ has LESS than $-oo$ leading $0s) = 0$ but it is hard to add it to the probabilities while staying clear.}\\
%2$^{len(h(x))}\times \frac{2H_c-1}{H_c}$ items into the sketch.\\

Note that the above analysis for scenario S4 made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are 0, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert without increasing the HLL cardinality estimate becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$. Of course, the adversary can relax its attack and allow some of the empty buckets to be hit. This increases the insertion rate at the cost of a small increase in the final cardinality estimate.

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. The blind adversary in scenario S2 has no way to detect that it is attacking an empty HLL and cannot adapt its strategy, so all buckets that are hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality estimate of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S4 is able to detect that all buckets are empty and adapt its attack accordingly. The adversary can now target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other trade-offs between work and the cardinality estimate are of course possible.

In both scenarios S2 and S4, we see that an adversary who has $t$ bits of flexibility in its inputs can expect to insert exponentially (in $t$) many inputs values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the cases where the HLL sketch has empty buckets in S2 or where the HLL sketch is empty to begin with in S2 and S4). The result is largely independent of the HLL parameters, but does depend on the extent to which the HLL sketch is already filled, via a factor $1-(2H_c)^{-1}$ and the number of empty buckets $r$, via a factor $1- \frac{r}{2^n}$.

@@

As a proof-of-concept, we implemented our attacks against the HLL implementation from \cite{clahll}. The attack code is available at \href{https://github.com/PizzaWhisperer/HLLVuln}{\textit{\url{https://github.com/PizzaWhisperer/HLLVuln}}}. For simplicity, our items are 4-character long ASCII strings, giving us up to 7,311,616 distinct items that can be inserted. We choose $h$ to be the 32-bit Murmur3 hash function from \cite{murmur3code} (note that the library allows the hash function to be specified; we pick Murmur3  for our attack and consider it reasonable that the adversary should know the hash function in use). 
% \kenny{Why do we get to choose $h$? Isn't it fixed by the specific implementation we are attacking?} \mathilde{The library has a place holder for a 32 bit hash function but lets the user choose it. This was one of my concerns and I was wondering if it would make sense to attack \cite{datasketch} instead (hardcoded $h$, but when we discussed it, you said that due to Kerckhoff's principle, this hash function should be public so it does not matter that much.} 
We create an HLL sketch using $2^n = 256$ buckets and initialise it with 1,000 random items representing honest users' data.

We then challenge the adversary to pick the largest possible number of items to add to the sketch from our set of  7,311,616 possible inputs while trying to keep the cardinality estimate as low as possible, either in the S2 or the S4 scenario. As a second experiment, and in order to get results that can be directly compared to the work of~\cite{hllvuln}, we mounted our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose what to insert from a set of 250,000 random items that are given to him. Finally, for completeness, we also run the attacks against HLL sketches that are initially empty. All attacks were run 30 times and the outcomes averaged. Our results are reported in Tables~\ref{table:tab1} and~\ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary, and the cardinality estimate after adding the adversary's items.

These empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the adversary is able to hide approximately half of the items in our starting set under S2, and around 83\% under S4. %( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$) 
Here $0.83 \approx (1-\frac{1}{2\cdot H_c})$ with $H_c\approx\frac{1000}{256}=3.9$. Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by equation (29) in~\cite{hll}, or 1, in respectively S2, S4.

To assess the efficiency of our attacks, we can compare our results from Table~\ref{table:tab2} with the attack presented in the Section 5.1 of \cite{hllvuln}. In that paper, the authors target the Redis~\cite{redishll} open-source implementation of HLL and perform the attack against an empty sketch. Given a set of 250,000 items, the adversary could choose ``74,390 distinct items and obtain an HLL estimate of only 15,780``~\cite{hllvuln}, achieving a five-fold reduction from the true cardinality. In comparison, the adversary following our strategy is able to add 125,000 distinct items while keeping the cardinality estimate at 367, providing a reduction by a factor of 340 instead of 5.

\begin{table}[h]
\caption{Attack results, averaged over 30 iterations.}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Card. Est.} & 1,000 & 0 & 1,005 & 0 \\ \hline
    \textbf{\# Items added} & 3,655,744 & 3,655,740 & 6,090,003 & 14,220 \\ \hline
    \textbf{Final Card. Est.} & 1'011 & 367 & 1,005 & 1 \\ \hline
\end{tabular}
\label{table:tab1}
\end{table}

\begin{table}[h]
\caption{Attack results, averaged over 30 iterations, in the setting of \cite{hllvuln}.}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Card. Est.} & 1,004 & 0 & 998 & 0 \\ \hline
    \textbf{\# Items added} & 124,940 & 125,085 & 204,669 & 501 \\ \hline
    \textbf{Final Card. Est.} & 1,058 & 367 & 998 & 1 \\ \hline
\end{tabular}
\label{table:tab2}
\end{table}

\section{Conclusion}\label{sec:conclusions}
The HyperLogLog algorithm is an elegant and efficient solution to the problem of estimating the cardinality of large sets. Its simple structure makes it easy to code and use, as shown by the growing number of available open-source implementations. Nonetheless, malicious users can manipulate the HLL cardinality estimate and thence break the security properties of systems relying on HLL. Our attacks are simple but powerful and should raise awareness of the limitations of HLL. Our analysis may assist software developers in understanding the risks they run when using HLL in adversarial settings.

A formal analysis of using salting as a countermeasure to our attacks remains open.  
%Future work could consist in finding mitigations that do not rely on salt. For example, would the imposed ID setting (like Facebook) solve the issue but still be convenient for the various HLL applications? 
Also left for future work is the task of extending our attacks to HLL variants such as HLL++~\cite{hllpratice} and sliding HLL~\cite{slidinghll}.
